Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-07 ms/main.tex
Add Ohad's graph and some todo to discuss with Yoav
update todos

With both vertical and oblique transmission, $0<v<1$, the recursion system is more complex
and we focus on local rather than on global stability.
To proceed,  note that 
\autoref{eq:horizontal} gives $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
\autoref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
\autoref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
%TODO cleanup 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}),
determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

@@

This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection.
Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), making it easier for cooperation to evolve.
However, increasing the horizontal transmission of defection can either increase or decrease the boundary, making it easier or harder for cooperation to evolve, respectively, depending if $T_A$ is high or low (i.e. if $T_A>1/(1+bv\alpha)$ or below).
%TODO I think the boundary on TA should be (2TB-1)/(alpha*b*v+1). DC
Increasing the vertical transmission $v$ can either increase or decrease the boundary since $\partial \hat\gamma_1 / \partial v = \frac{(1-T_B)(T_B-T_a)}{v^2(1-T_B)^2}$ which can be both positive and negative depending on the sign of $T_B-T_A$.
When $T_B>T_A$ we get $\partial \hat\gamma_1 / \partial v >0$ and therefore, as the vertical transmission increases the boundary on the cost increases making it easier for cooperation to evolve from rarity.
Similarly, when $T_A > T_B$ we get $\partial \hat\gamma_1 / \partial v <0$ and therefore, as the vertical transmission increases the boundary on the cost decreases making it harder for cooperation to evolve from rarity.
We got that when $T_A>T_B$ vertical transmission is interfering cooperation to evolve by decreasing the maximum cost that the cooperates can pay and stil evolve from rairty. 
%TODO Is the last sentence requried ? Should we mention here we already saw result 1 - when $T_A > T_B$ we will have global fixation of cooperation

%($\text{sign}\big(\partial \hat\gamma_1 / \partial T_B \big)= \text{sign}\big(b v \alpha T_A - (1-T_A)\big)$).
%TODO talk about partial gamma 1 / partial v = -(TA-TB)(1-TB)
%TODO - do we need the command above ? DC

@@

Note that $\partial \hat a_2 / \partial v = b T_A (T_A-T_B)$. $\partial \hat a_2 / \partial v > 0$ when $T_A > T_B$ and therefore, as the vertical transmission $v$ \emph{increases} a greater value of social association is required for fixation of cooperation from rarity.
Similarly, when $T_B>T_A$ $T_A > T_B$ we get $\partial \hat a_2 / \partial v < 0$ and therefore, as $v$ \emph{decreases} a greater value of social association is required for fixation of cooperation from rarity.
\\

\begin{corollary}[Necessary condition for fixation of cooperation]
\label{corollary:vhat}
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
Note that this condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{With population structure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Ayelet Shavit for discussions and comments.
This work was supported in part by
the Israel Science Foundation 552/19 (YR),
and Minerva Stiftung Center for Lab Evolution (YR).
% TODO funding for other authors?
}

@@

%%% Figure: Spatial model
\begin{figure}[h]
  \centering
%   \includegraphics[scale=0.5]{spatial1.png}
    ~
 %   \includegraphics[scale=0.5]{spatial2.png} \\
  \caption{
  \textbf{Evolution of cooperation in a spatial model.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of the left panel, or the transmission bias $T_A-T_B$ on the x-axis of the right panel.
  The population evolves on a 100-by-100 grid. Cooperation and horizontal cultural transmission are all local between adjacent sites, and each site had 8 neighbors.
  The black curves represent the cost boundaries for the evolution of cooperation in a well-mixed population with social association where $\alpha=1/8$ in (left) \autoref{eq:equal_transmission} and (right) \label{eq:cost_boundaries}.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed and 50 simulations were executed for each parameter set.
  Here, population size is $10,000$ (100-by-100 grid), benefit of cooperation, $b=1.3$, perfect vertical transmission $v=1$.
  \textbf{(left)} Symmetric horizontal transmission, $T=T_A=T_B$. \textbf{(right)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$.
  }
  \label{fig:spatial}
\end{figure}

>>>>

With both vertical and oblique transmission, $0<v<1$, the recursion system is more complex
and we focus on local rather than on global stability.
To proceed,  note that 
\autoref{eq:horizontal} gives $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
\autoref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
\autoref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
%TODO YR cleanup  
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}),
determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

@@

This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection.
Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), making it easier for cooperation to evolve.
However, increasing the horizontal transmission of defection can either increase or decrease the boundary, making it easier or harder for cooperation to evolve, respectively, depending if $T_A$ is high or low (i.e. if $T_A>1/(1+bv\alpha)$ or below).
% TODO write sign(dgamma1/dTB) = sign(???)
%TODO I think the boundary on TA should be (2TB-1)/(alpha*b*v+1). DC

Increasing the vertical transmission $v$ can either increase or decrease the boundary, depending on the horizontal transmission bias, $T_A-T_B$ (because $\text{sign}(\partial \hat\gamma_1 / \partial v) = -\text{sign}(T_A-T_B)$.)
When $T_A<T_B$ we get $\partial \hat\gamma_1 / \partial v >0$ and therefore, as the vertical transmission increases the boundary on the cost increases making it easier for cooperation to evolve from rarity. 
Similarly, when $T_A > T_B$ we get $\partial \hat\gamma_1 / \partial v <0$ and therefore, as the vertical transmission increases the boundary on the cost decreases making it harder for cooperation to evolve from rarity.
We got that when $T_A>T_B$ vertical transmission is interfering cooperation to evolve by decreasing the maximum cost that the cooperates can pay and stil evolve from rairty. 
% TODO YR revise; note that without vertical transmission TA-TB determines the outcome

@@

Note that $\partial \hat a_2 / \partial v = b T_A (T_A-T_B)$. $\partial \hat a_2 / \partial v > 0$ when $T_A > T_B$ and therefore, as the vertical transmission $v$ \emph{increases} a greater value of social association is required for fixation of cooperation from rarity. %TODO YR revise
Similarly, when $T_B>T_A$ $T_A > T_B$ we get $\partial \hat a_2 / \partial v < 0$ and therefore, as $v$ \emph{decreases} a greater value of social association is required for fixation of cooperation from rarity.
\\

\begin{corollary}[Necessary condition for fixation of cooperation]
\label{corollary:vhat}
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
This condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{With Population Structure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany, Ayelet Shavit, and Kaleda Krebs Denton for discussions and comments.
This work was supported in part by
the Israel Science Foundation 552/19 (YR),
and Minerva Stiftung Center for Lab Evolution (YR).
% TODO funding for other authors?
}

@@

%%% Figure: Spatial model
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.5]{{Ta=Tb_b=1.3_globalSelection}.pdf}
    ~
    \includegraphics[scale=0.5]{{Ta=0.4_Tb=0.3-0.5_b=1.3_globalSelection}.pdf} \\
  \caption{
  \textbf{Evolution of cooperation in a spatial model.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of the left panel, or the transmission bias $T_A-T_B$ on the x-axis of the right panel.
  The population evolves on a 100-by-100 grid.
  Cooperation and horizontal transmission are both local between adjacent sites, and each site had 8 neighbors.
  Selection operates globally (see Fig X for results from a model with local selection).
  The black curves represent the cost boundaries for the evolution of cooperation in a well-mixed population with social association where $\alpha=1/8$ in (left) \autoref{eq:equal_transmission} and (right) \label{eq:cost_boundaries}.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed and 50 simulations were executed for each parameter set.
  Here, population size is 10,000 (100-by-100 grid), benefit of cooperation, $b=1.3$, perfect vertical transmission $v=1$.
  \textbf{(left)} Symmetric horizontal transmission, $T=T_A=T_B$.
  \textbf{(right)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$.
  }
  \label{fig:spatial}
\end{figure}

% TODO add figure of fluctuations in spatial model
% TODO add figure of local selection in spatial model
